Split-Apply-Combine with Dynamic Grouping

### 2023-03-14 paper/article.Rnw
wip

In the simplest case, where no auxiliary information is available or used, one
replaces the estimator of the mean over a subpopulation with the estimator of
the mean over a larger subpopulation that includes the target subpopulation
\citep[Section 3.2.1]{rao2015small}. If we assume that the dataset is obtained
by simple random sampling from the population, the mean can be estimated with
the sample mean. In this example we will demand that there are at least ten
records for which turnover has been measured. The collapsing scheme is given by
\code{sbi*size ~ sbi + sbi2 + sbi1} where \code{sbi2} and \code{sbi1} are
classification by respectively the first two SBI digits and the first SBI
digit. We first add those variables to the dataset.
<<>>=
producers$sbi2 <- substr(producers$sbi,1,2)
producers$sbi1 <- substr(producers$sbi,1,1)
head(producers, 3)
@
Using \code{accumulate()} we obtain the means. 
<<>>=
a <- cumulate(producers
             , collapse = sbi*size ~ sbi + sbi2 + sbi1
             , test = min_complete(n=10, vars="industrial")
             , mean_industrial = mean(industrial, na.rm=TRUE))
head(a,3)
@

@@

\subsection{Imputing missing values using SAE and ratio imputation} 
Our goal in this example is to impute missing values for the \code{industrial}
variable based on ratio imputation with \code{total} as predictor.  Ratio
imputation is a method where the imputed value $\hat{y}_i$ for variable $Y$ of
record $i$ is estimated as $\hat{y_i}=\hat{R}_d x_i$, where $\hat{R}$ is an
estimate of the ratio between $Y$ and an auxiliary variable $X$ in
subpopulation $d$. An unbiased estimate for $R_d$ is given by
$\hat{\bar{Y}}_d/\hat{\bar{X}}_d$ where $\hat{\bar{Y}}_d$ and $\hat{\bar{X}}_d$
are estimated subpopulation means.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computing complex aggregates} \label{sect:complex}
Generalizing further, one may be interested in estimating (statistical) models.
With the \code{cumulate} function it is possible to specify complex aggregates
that go beyond a simple number. Below, we estimate the following linear model
\begin{displaymath}
\texttt{total} = \beta_0 + \beta_i\texttt{industrial} + \varepsilon,
\end{displaymath}
demanding that there are at least 10 records where both predictor and predicted
variable are available.
<<>>=
r <- cumulate(producers
         , collapse = sbi*size ~ sbi + sbi2 + sbi1
         , test = min_complete(n=10, vars=c("total","industrial"))
         , model = lm(total ~ industrial))
head(r,3)
@
Here, the last column is a list of class \code{object_list}, where each
element is either an object of class \code{lm} or \code{NA}. Variables of class
\code{object_list} only differ from from a standard \proglang{R} list by their
print method.

>>>>

In the simplest case, where no auxiliary information is available or used, one
replaces the estimator of the mean over a subpopulation with the estimator of
the mean over a larger subpopulation that includes the target subpopulation
\citep[Section 3.2.1]{rao2015small}. If we assume that the dataset is obtained
by simple random sampling from the population, the mean can be estimated with
the sample mean. In this example we will demand that there are at least ten
records for which turnover has been measured. The collapsing scheme is given by
\code{sbi*size ~ sbi + sbi2 + sbi1} where \code{sbi2} and \code{sbi1} are
classification by respectively the first two SBI digits and the first SBI
digit. We first add those variables to the dataset.
<<>>=
producers$sbi2 <- substr(producers$sbi,1,2)
producers$sbi1 <- substr(producers$sbi,1,1)
head(producers, 3)
@
Using \code{cumulate()} we obtain the means. 
<<>>=
a <- cumulate(producers
             , collapse = sbi*size ~ sbi + sbi2 + sbi1
             , test = min_complete(n=10, vars="industrial")
             , mean_industrial = mean(industrial, na.rm=TRUE))
head(a,3)
@

@@

\subsection{Imputing missing values using SAE and ratio imputation} \label{sect:ratio}
Our goal in this example is to impute missing values for the \code{industrial}
variable based on ratio imputation with \code{total} as predictor.  Ratio
imputation is a method where the imputed value $\hat{y}_i$ for variable $Y$ of
record $i$ is estimated as $\hat{y_i}=\hat{R}_d x_i$, where $\hat{R}$ is an
estimate of the ratio between $Y$ and an auxiliary variable $X$ in
subpopulation $d$. An unbiased estimate for $R_d$ is given by
$\hat{\bar{Y}}_d/\hat{\bar{X}}_d$ where $\hat{\bar{Y}}_d$ and $\hat{\bar{X}}_d$
are estimated subpopulation means.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Random Nearest neighbours imputation with collapsing groups}
Nearest neighbor (NN) imputation is a donor imputation method where the
imputation value is copied from a (possibly random) donor record with
attributes that are sufficiently similar \citep{andridge2010review}.
One way to generate an imputation value is to select a random donor
from the same group. To prevent the same donor from being used too
often, it is not uncommon to demand a minimum number of records
in the donor pool. A collapsing scheme is one way of guaranteeing
this, and below we demonstrate how this problem can be expressed
in \code{accumulate}.

We wish to impute the variable \code{trade} in the \code{producers} dataset
using donor imputation, where donors come from the same (\code{sbi},
\code{size}) combination. We wish to sample donor values from a group of at
least 5 donors. If this is not possible, we use the same fallback scenario as
in the previous Section.

We first define an `aggregation' function that takes a vector 
of donors and returns a non-empty sample.
<<>>=
random_element <- function(x) sample(x[!is.na(x)], 1)
@
To make sure we obtain a donor for each record, we add an
identifying column to use as grouping variable.
\code{id} column.
<<>>=
producers <- cbind(id = sprintf("ID%03d",seq_len(nrow(producers)))
                  , producers)
set.seed(111)
imputations <- cumulate(producers
                 , collapse = id ~ sbi*size + sbi + sbi2 + sbi1
                 , test = min_complete(5,"trade")
                 , donor_trade = random_element(trade))

head(imputations, 3)
@
To use the donor imputations, we merge the imputation candidates with the
original dataset and use \code{impute_proxy()} of the \pkg{simputation} package
for imputation.
<<>>=
imputed <- merge(producers, imputations) |>
           impute_proxy(trade ~ donor_trade)
cols <- c(1:3,9:10,5)
head(producers[cols],3)
head(imputed[c(cols,11)],3)
@
The merge operation automatically merges on the \code{id} column, which also
adds the \code{level} column to the output.  The function \code{impute_proxy}
copies values from \code{donor_trade} into trade where \code{trade} is missing.
In the last expressions we only print the columns of interest.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computing complex aggregates} \label{sect:complex}
Until now, the aggregates have been simple (scalar) values.  With the
\code{cumulate()} function it is also possible to specify complex aggregates that
go beyond simple aggregates. Below, we estimate the following linear model
\begin{displaymath}
\texttt{total} = \beta_0 + \beta_i\texttt{industrial} + \varepsilon,
\end{displaymath}
demanding that there are at least 10 records where both predictor and predicted
variable are available.
<<>>=
r <- cumulate(producers
         , collapse = sbi*size ~ sbi + sbi2 + sbi1
         , test = min_complete(n=10, vars=c("total","industrial"))
         , model = lm(total ~ industrial))
head(r,3)
@
Here, the last column is a list of class \code{object_list}, where each
element is either an object of class \code{lm} or \code{NA}. Variables of class
\code{object_list} only differ from from a standard \proglang{R} list by their
print method.
